### 2022-07-09 ehoh/ehoh.tex
geoff ch 3

Our vehicle is E\begin{rep} \cite{ss-02-brainiac,scv-19-e23}\end{rep},
a prover developed primarily by Schulz. It is written in C and offers
good performance, with more emphasis on ``brainiac''
heuristics than on raw speed. E regularly scores among the top
systems at CASC and is usually the strongest open-source prover in
the relevant divisions. It also serves as a backend for
competitive higher-order provers. We refer to our extended version of
E as Ehoh. It corresponds to a prerelease version of E~2.5 configured with the
option \verb|--enable-ho|.%
\footnote{\url{https://github.com/eprover/eprover/commit/80946ac}}
%We plan to include the new higher-order inferences in the E 2.5 release.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free
higher-order logic (\lfhol{}) described by Bentkamp et al.\
%, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
and its semantics by interpreting the Boolean type $o$ as a domain of
cardinality~2. Functional extensionality, the property that two functions are equal if they
always return the same value when given the same arguments, can be obtained by
adding suitable axioms
\cite[Sect.~3.1]{bbcw-21-lfho}.

@@

\looseness=-1
A side effect of the flattened representation is that prefix subterms are not
shared. For example, the terms $\cst{f}\;\cst{a}$ and $\cst{f}\;\cst{a}\;\cst{b}$
correspond to the flattened cells $\cst{f}(\cst{a})$ and
$\cst{f}(\cst{a}, \cst{b}).$ The argument subterm $\cst{a}$ is shared, but not
the prefix $\cst{f}\;\cst{a}.$ Similarly,
$x$ and $x\;\cst{b}$ are represented by two distinct cells, $x()$ and
$x(\cst{b})$, and there is no connection between the two occurrences of~$x$.
%
In particular,
despite perfect sharing, their \verb|binding| fields are unconnected, leading
to inconsistencies.

@@

\ourpara{Term Orders} Superposition provers rely on term orders to prune the search
space. The order must be a reduction order
%can be extended to a simplification order that
% That's true but pointless. It's equivalent to saying that you need a
% simplification order and the prover may underapproximate it. -JB
(Sect.~\ref{sec:pre:order}). E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
(LPO). KBO is widely regarded as the more
robust option for superposition.
%
In earlier work, Blanchette and colleagues have shown that only KBO can be
generalized gracefully while preserving the necessary properties for
superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
reason, we focus on KBO.

E implements L\"ochner's linear-time algorithm for KBO
\cite{bl-06-kbo}, which relies on the \relax{tupling} method to store
intermediate results. % , avoiding repeated computations.
It is straightforward to generalize the algorithm to compute the
graceful \lfhol{} version of KBO \cite{bbww-17-kbo}.
The main difference is that when comparing two terms $\cst{f} \; \tuple{s}{m}$
and $\cst{f} \; \tuple{t}{n}$, because of partial application we may now
have $m \not= n$; this required changing the implementation to
perform a length-lexicographic comparison of the tuples $\tuple{s}{m}$ and
$\tuple{t}{n}.$

@@

  A unification algorithm for \lfhol{} can be derived from the above transition
  system, by committing to a strategy for applying the rules.
  %
  This algorithm closely follows the Ehoh implementation, abstracting away from
  complications such as prefix optimization. We assume a flattened
  representation of terms; as in Ehoh, each variable stores the term it is bound
  to in its \textit{binding} field (Sect.~\ref{sec:ehoh:types-and-terms}).
  We also rely on a \textsc{ApplySubst} function, which applies the
  binding to the top-level variable. The algorithm assumes that the terms
  to be unified have the same type. The pseudocode is as follows:

@@

Given $s$ and $t$, the matching problem consists of finding a
substitution $\sigma$ such that $\sigma(s) = t.$
We then write that ``$t$ is an instance of $s$'' or
``$s$~generalizes~$t$.'' %and call $\sigma$ a \emph{generalizing substitution}.
We are interested in most general generalizations (MGGs).
Matching can be reduced to unification by treating variables in $t$ as
nullary symbols \cite{bn-98-tr-and-all-that}, but E implements it
separately.

@@

\looseness=-1
When generalizing the data structure to \lfhol, we face a complication
due to partial application. First-order terms can only be stored in leaf nodes,
but in Ehoh we must also be able to represent partially applied terms,
such as $\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that
$\cst{f}$ is unary and $\cst{g}$ is binary). Conceptually, this can be solved
by storing a Boolean on each node indicating whether it is an accepting state.
In the implementation, the change is more subtle, because several parts of E's
code implicitly assume that only leaf nodes are accepting.

@@

  The core functionality is implemented in \textsc{FindNext\-Node}, presented below. This
  procedure moves the iterator to the next node that has not been explored in the
  search for generalization, or \textit{Null} if the entire tree has been
  traversed. It first goes through all child nodes labeled with a variable before
  possibly visiting the child node labeled with a function symbol. We assume that
  we can iterate through the children of a node using a function
  \textsc{NextVarChild} that, given a tree node and iterator through children,
  advances the iterator to the child representing the next variable. Furthermore, we
  assume that the iterator can also be in the distinguished states \textit{Start}
  and \textit{End}.
  \textit{Start} indicates that no child has been visited yet; \textit{End}
  indicates that we have visited all children. Finally, the expression
  $n.\mathit{child}(\zeta)$ returns a child of the node $n$ labeled $\zeta$ if
  such a child exists or $\mathit{Null}$ otherwise.

@@

The pseudocode uses a slightly different representation of backtracking tuples
than ${\Pdtarrow}$. In the \textsf{AdvanceX} rule,
$\sigma$ changes only if the variable~$x$ was previously not bound.
Instead of creating and storing substitutions explicitly, we simply remember
whether the variable was bound in this step or not, in the
$\mathit{var\_unbound}$ tuple component. Then we rely on the label~$x$ of the
current node and its \textit{binding} field to carry the substitutions.
Similarly, since our strategy is to traverse the tree by first visiting the
variable-labeled child nodes, we need to remember how far we have come with
this traversal. We store this information in the $\mathit{c\_iter}$ tuple
component.

@@

Since Ehoh is based on a mono\-morphic logic, the only way to support
extensionality without changing the calculus is to add a set of extensionality
axioms for every function type occurring in the problem
\cite[Sect.~3.1]{bbcw-21-lfho}. The evaluation by Bentkamp et al.\ of such
an approach was discouraging \cite[Sect.~6]{bbcw-21-lfho}, so we decided to
support extensionality via inference rules in Ehoh. We implemented two well-known
incomplete rules we had experimented with in the context of Zipperposition.

@@

E's heuristics are largely independent of the logic used and work
unchanged for Ehoh. Yet, in preliminary experiments, we noticed that E proved
some \lfhol{} benchmarks quickly using the applicative encoding
(Sect.~\ref{sec:ehoh:introduction}), whereas Ehoh timed out.
\begin{rep}There were enough such problems to prompt us to take a
closer look. \end{rep}%
Based on these observations, we extended the heuristics\confrep{}{ to exploit
\lfhol{}-specific features}.

@@

In preliminary experiments, we discovered that the presence of an explicit
application operator \appvar{} can be beneficial for some problems.
\begin{rep}
%A small example will help illustrate this behavior.
Let $\cst{a} : \iota_1$,
$\cst{b} : \iota_2$, $\cst{c} : \iota_3$, $\cst{f} : \iota_1 \to \iota_2 \to
\iota_3$, $x : \iota_2 \to \iota_3$, $y : \iota_2$, and $z : \iota_3$, and
consider the clauses
%
$\cst{f} \; \cst{a} \; y \noteq \cst{c}$ and
$x \; \cst{b} \eq z$,
%
where the first one is the negated conjecture. Their applicative encoding
is
%
$\cst{@}_{\iota_2,\iota_3}(\cst{@}_{\iota_1,\iota_2\to\iota_3}(\cst{f}, \cst{a}), y) \noteq \cst{c}$ and
$\cst{@}_{\iota_2,\iota_3}(x, \cst{b}) \eq z$,
%
where $\cst{@}_{\tau,\upsilon}$ is a type-indexed family of
symbols representing the application of a function of type $\tau
\to \upsilon$.
\end{rep}%
With the applicative encoding, generation schemes can take the symbols
$\cst{@}_{\tau,\upsilon}$ into account, thereby exploiting the type
information carried by such symbols.\begin{rep} Since $\cst{@}_{\iota_2,\iota_3}$ is a
conjecture symbol, some weight generation scheme could give it a low weight,
which would also impact the second clause. By contrast, the native \lfhol{}
clauses share no symbols; the connection between them is hidden in the types
of variables and symbols, which are ignored by the heuristics.

@@

The side conditions of the superposition rules SN and SP
(Sect.~\ref{sec:pre:rules}) rely on a literal
selection function to restrict the set of \emph{inference literals},
thereby reducing the search space. Given a clause, a literal selection
function returns a (possibly empty) subset of its literals. For
completeness, any nonempty subset selected must contain at least
one negative literal. If no literal is selected, all \emph{maximal}
literals become inference literals.
%Given a clause $L_1 \lor \cdots \lor L_n$, a literal selection function
%returns either $\{L_i\}$ for some $i$ or $\emptyset$.
The most widely used function is probably
\texttt{SelectMaxLComplexAvoidPosPred}, which we abbreviate to
\texttt{SelectMLCAPP}. It selects at most one negative literal, based
on size, absence of variables, and maximality of the literal in the clause.
%It
%also avoids negative literals that share a predicate symbol with a
%positive literal in the same clause.

\begin{rep}
  Intuitively, applied variables can potentially be unified with more terms than
  terms with rigid heads. This makes them prolific in terms of possible
  inference partners, a behavior we might want to avoid. On the other hand, shorter proofs
  might be found if we prefer selecting applied variables. To cover both
  scenarios, we implemented selection functions that prefer or defer selecting
  applied variables. % and that break ties using \texttt{SelectMLCAPP}.

@@

  Selection of the given clause is a critical choice point.
  %To ensure completeness every candidate clause should be selected
  %eventually\confrep{}{, unless it is identified as redundant and
  %discarded before its turn has come}. Ignoring a candidate forever
  %would compromise E's completeness guarantee.
  E heuristically assigns \emph{clause priorities} and \emph{clause weights} to
  the candidates.
  The priorities provide a crude partition, whereas the weights order the clauses
  within a partition.
  E's main loop visits, in round-robin fashion, a set of priority queues. From a given
  queue, the clause with the highest priority and the smallest weight is selected.
  \begin{rep}%
  Typically, one of the queues will use the clauses' age as priority, to ensure
  fairness.
  \end{rep}

@@

}
Configurations that are suitable for a wide range of problems have emerged over
time. One of them is the configuration that is most often chosen by E's \emph{auto}
mode. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
%
%%% TYPESETTING: TEMPORARILY DISABLED LEFTMARGIN (?)
\begin{description}[labelwidth=\widthof{Precedence generation:}]
  \item[\rm Term order:] KBO
  \item[\rm Weight generation:] \texttt{invfreqrank}
  \item[\rm Precedence generation:] \texttt{invfreq}
  \item[\rm Literal selection:] \texttt{SelectMLCAPP}
  \item[\rm Clause selection:] ~ \\
\verb|1.CRSWeight(SimulateSOS, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.CRSWeight(ConstPrio, 0.1, 100, 100, 100, 100, 1.5, 1.5, 1.5),| \\
\verb|1.FIFOWeight(PreferProcessed),| \\
\verb|1.CRSWeight(PreferNonGoals, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.Refinedweight(SimulateSOS, 3, 2, 2, 1.5, 2)|
\end{description}
%
The clause selection scheme consists of five queues, each of which is
specified by a weight function template. The prefixes $n$\texttt{.}\ next to
the template names indicate that the queue will be visited $n$ times in the round-robin scheme
before moving to the next one. The first argument to each
template is the clause priority scheme.

@@

For first-order logic, skolemization preserves both satisfiability
(unprovability) and unsatisfiability (provability). In contrast, for
higher-order logics without the axiom of choice, naive skolemization is
unsound, because it introduces symbols that can be used to instantiate
higher-order variables.
%
One solution proposed by Miller \cite[Sect.~6]{dm-87-compact-proofs} is to ensure that
Skolem symbols are always applied to a minimum number of arguments.
%Accordingly, the version of \lfhol{} introduced by Bentkamp et al.\
%\cite{bbcw-21-lfho} distinguishes between mandatory and optional
%arguments.
However, to keep the implementation simple, we have decided to
ignore this issue and consider all arguments as optional, including those to
Skolem symbols. In Chapter \ref{ch:ehoh2} we extend Ehoh's logic to full higher-order logic
with the axiom of choice, which addresses the issue.
%TODO_PHD: Once Ehoh 2 paper is done, add reference to the section about choice here.

There is another transformation performed by preprocessing that is
problematic, but for a different reason. \emph{Definition unfolding} is the
process of replacing equationally defined symbols with their definitions and
removing the defining equations. A \relax{definition} is a clause of the form
$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
\{x_1,\dots,x_m\}$.
This transformation preserves unsatisfiability (provability) for first-~and
higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
is that by removing the definitional clause, we also remove a symbol $\cst{f}$
that otherwise could be used to instantiate a higher-order quantifier.
For example, the clause set
$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
higher-order logic, the second clause set would be unsatisfiable thanks to
the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-reduction.)
%
For the moment, we have simply disabled definition unfolding in Ehoh. We have
not measured the effect of this choice, but we conjecture it is not substantial.
Namely, appropriately chosen term order will result in a similar outcome without
removing the defined symbol. We will enable definition unfolding again once we
have added support for $\lambda$-terms.

@@

The output may contain $\itrue$, $\ifalse$, or Bool\-ean
variables as nested formulas. Since E was first developed as an untyped prover,
unification of a variable with a Boolean constant was disallowed to avoid
unsoundness. We needed to undo this in Ehoh. Ehoh must also remove trivial
literals $\ifalse \eq \itrue$ and $\itrue \noteq
\itrue$ that emerge during proof search.
\end{rep}

@@

%In this section, we consider the following questions:
How useful are Ehoh's new heuristics? And how does Ehoh perform compared with
E, used directly or in tandem with the
applicative encoding, and compared with other provers?
To answer the first question, we evaluated each new parameter independently. From
the empirical results, we derived a new configuration optimized for \lfhol{}.
For the second question, we compared Ehoh's success
rate\begin{rep} and speed\end{rep} on \lfhol{} problems with
native higher-order provers and
on applicatively encoded problems with E. We also included first-order
benchmarks to measure Ehoh's overhead.

@@

We used the \emph{boa} configuration as the basis to evaluate the new
heuristic schemes. For each heuristic parameter we tuned, we changed only its
value while keeping the other parameters the same as for \emph{boa}.
\begin{rep}This gives an idea of how each parameter affects overall
performance.\end{rep}
All heuristic parameters were tested on a 5012~problem suite generated
using Sledgehammer, consisting of four variants of the Judgment Day
\cite{bn-10-sh} suite.\begin{rep} The problems were given in native
\lfhol{} syntax.\end{rep} The experiments described in this subsection were
carried out using an earlier E version (2.3).\begin{conf}
%, exploiting its support for partial application and applied variables.
Our main findings are as follows:
%
\begin{itemize}
\item The combination of the weight generation scheme \texttt{invtypefreqrank}
  and the precedence generation scheme \texttt{invtypefreq} performs best.

@@

Clause selection is the heuristic component that we extended the most. We must
assess the effect of a new heuristic weight function, a multiplier for the
occurrence of top-level applied variables, and clause priority based on the
number of top-level applied variables.

@@

To test the effect of the new type-based weight function, we replaced
\emph{boa}'s queue, which uses
\texttt{4.CRS\-Weight(}\ldots\texttt{)}, with the queue ordered by
\texttt{4.CRSTypeWeight(}\ldots\texttt{)}. We call the original
heuristic \texttt{W} and the type-aware alternative \texttt{TW}.
%
We chose nine values for testing the effect of the applied
variable multiplier \textit{appv\_mult}.
%
Figure \ref{fig:ehoh:clauseweight-fun-eval} summarizes the results of combining
\texttt{W} or \texttt{TW} with the different
\textit{appv\_\allowbreak mult} values. Applying a multiplier smaller than
1, which corresponds to preferring literals containing applied variables, can
lose dozens of solutions. Overall, using the type-aware heuristic seems
slightly detrimental.

>>>>

Our vehicle is E\begin{rep} \cite{ss-02-brainiac,scv-19-e23}\end{rep},
a prover developed primarily by Schulz. It is written in C and offers
good performance. It gets its ``brainiac'' name by emphasizing 
intelligent heuristics more than raw speed. E regularly scores among the top
systems at CASC and is usually the strongest open-source prover in
the relevant divisions. It also serves as a backend for
competitive higher-order provers. We refer to our extended version of
E as Ehoh. It corresponds to a prerelease version of E~2.5 configured with the
option \verb|--enable-ho|.%
\footnote{\url{https://github.com/eprover/eprover/commit/80946ac}}
%We plan to include the new higher-order inferences in the E 2.5 release.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free
higher-order logic (\lfhol{}) described by Bentkamp et al.\
%, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
and its semantics by interpreting the Boolean type $o$ as a two-element domain. Functional extensionality, the property that two functions are equal if they
always return the same value when given the same arguments, can be obtained by
adding suitable axioms
\cite[Sect.~3.1]{bbcw-21-lfho}.

@@

\looseness=-1
A side effect of the flattened representation is that prefix subterms are not
shared. For example, the terms $\cst{f}\;\cst{a}$ and $\cst{f}\;\cst{a}\;\cst{b}$
correspond to the flattened cells $\cst{f}(\cst{a})$ and
$\cst{f}(\cst{a}, \cst{b}).$ The argument subterm $\cst{a}$ is shared, but not
the prefix $\cst{f}\;\cst{a}.$ Similarly,
$x$ and $x\;\cst{b}$ are represented by two distinct cells, $x()$ and
$x(\cst{b})$, and there is no connection between the two occurrences of~$x$.
%
In particular, when $x()$'s \verb|binding| field is updated, this does not affect the \verb|binding| of $x(\cst{b})$.

@@

\ourpara{Term Orders} Superposition provers rely on term orders to prune the search
space. The order must be a reduction order
%can be extended to a simplification order that
% That's true but pointless. It's equivalent to saying that you need a
% simplification order and the prover may underapproximate it. -JB
(Sect.~\ref{sec:pre:order}). E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
(LPO). KBO is widely regarded as the more
robust option for superposition.
%
In earlier work, Blanchette and colleagues have shown that only KBO can be
generalized gracefully while preserving the necessary properties for
superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
reason, in this chapter, we focus on KBO.

E implements L\"ochner's linear-time algorithm for KBO
\cite{bl-06-kbo}, which relies on the \relax{tupling} method to store
intermediate results. % , avoiding repeated computations.
It is straightforward to generalize the algorithm to compute the
graceful \lfhol{} version of KBO \cite{bbww-17-kbo}.
The main difference is that when comparing two terms $\cst{f} \; \tuple{s}{m}$
and $\cst{f} \; \tuple{t}{n}$, because of partial application it is possible that $m \not= n$; 
this required changing the implementation to
perform a length-lexicographic comparison of the tuples $\tuple{s}{m}$ and
$\tuple{t}{n}.$

@@

  A unification algorithm for \lfhol{} can be derived from the above transition
  system, by committing to a strategy for applying the rules.
  %
  An algorithm which closely follows the Ehoh implementation, abstracting away from
  complications such as prefix optimization is presented below. We assume a flattened
  representation of terms; as in Ehoh, each variable stores the term it is bound
  to in its \textit{binding} field (Sect.~\ref{sec:ehoh:types-and-terms}).
  We also rely on a \textsc{ApplySubst} function, which applies the
  binding to the top-level variable. The algorithm assumes that the terms
  to be unified have the same type. The pseudocode is as follows:

@@

Given $s$ and $t$, the matching problem consists of finding a
substitution $\sigma$ such that $\sigma(s) = t.$
We then write that ``$t$ is an instance of $s$'' or
``$s$~generalizes~$t$.'' %and call $\sigma$ a \emph{generalizing substitution}.
We are interested in most general generalizations (MGGs).
Matching can be reduced to unification by treating variables in $t$ as
nullary symbols \cite{bn-98-tr-and-all-that}, but E implements matching
separately.

@@

\looseness=-1
When generalizing the data structure to \lfhol, we face a complication due to
partial application. First-order terms can only be stored in leaf nodes, but in
Ehoh we must also be able to represent partially applied terms, such as
$\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that $\cst{f}$
is unary and $\cst{g}$ is binary). Conceptually, this can be solved by storing a
Boolean on each node indicating whether it is an accepting state (i.e., if it
corresponds to a term in the indexed set). In the implementation, the change is
more subtle, because several parts of E's code implicitly assume that only leaf
nodes are accepting.

@@

  The core functionality is implemented in \textsc{FindNext\-Node}, presented below. This
  procedure moves the iterator to the next node that has not been explored in the
  search for generalization, or \textit{Null} if the entire tree has been
  traversed. It first goes through all child nodes labeled with a variable before
  possibly visiting the child node labeled with a function symbol. The children of a node can be iterated through using a function
  \textsc{NextVarChild} that, given a tree node and iterator through children,
  advances the iterator to the child representing the next variable. Furthermore, we
  assume that the iterator can also be in the distinguished states \textit{Start}
  and \textit{End}.
  \textit{Start} indicates that no child has been visited yet; \textit{End}
  indicates that we have visited all children. Finally, the expression
  $n.\mathit{child}(\zeta)$ returns a child of the node $n$ labeled $\zeta$ if
  such a child exists or $\mathit{Null}$ otherwise.

@@

The pseudocode uses a slightly different representation of backtracking tuples
than ${\Pdtarrow}$. In the \textsf{AdvanceX} rule,
$\sigma$ changes only if the variable~$x$ was previously not bound.
Instead of creating and storing substitutions explicitly, the algorithm simply remembers
whether the variable was bound in this step or not, in the
$\mathit{var\_unbound}$ tuple component. Then it relies on the label~$x$ of the
current node and its \textit{binding} field to carry the substitutions.
Similarly, since the strategy is to traverse the tree by first visiting the
variable-labeled child nodes, it needs to remember how far it has come with
this traversal. This information is stored in the $\mathit{c\_iter}$ tuple
component.

@@

Since Ehoh is based on a mono\-morphic logic, the only way to support
extensionality without changing the calculus is to add a set of extensionality
axioms for every function type occurring in the problem
\cite[Sect.~3.1]{bbcw-21-lfho}. The evaluation by Bentkamp et al.\ of such
an approach was discouraging \cite[Sect.~6]{bbcw-21-lfho}, so we decided to
support extensionality via inference rules in Ehoh. We implemented two well-known
incomplete rules my colleagues and me had experimented with in the context of Zipperposition.

@@

E's heuristics are largely independent of the logic used, and work
unchanged for Ehoh. Yet, in preliminary experiments, we noticed that E proved
some \lfhol{} benchmarks quickly using the applicative encoding
(Sect.~\ref{sec:ehoh:introduction}), whereas Ehoh timed out.
\begin{rep}There were enough such problems to prompt us to take a
closer look. \end{rep}%
Based on these observations, we extended the heuristics\confrep{}{ to exploit
\lfhol{}-specific features}.

@@

In preliminary experiments, we discovered that the presence of an explicit
application operator \appvar{} can be beneficial for some problems.
\begin{rep}
%A small example will help illustrate this behavior.
Let $\cst{a} : \iota_1$,
$\cst{b} : \iota_2$, $\cst{c} : \iota_3$, $\cst{f} : \iota_1 \to \iota_2 \to
\iota_3$, $x : \iota_2 \to \iota_3$, $y : \iota_2$, and $z : \iota_3$, and
consider the clauses
%
$\cst{f} \; \cst{a} \; y \noteq \cst{c}$ and
$x \; \cst{b} \eq z$,
%
where the first one is the negated conjecture. Their applicative encoding
is
%
$\cst{@}_{\iota_2,\iota_3}(\cst{@}_{\iota_1,\iota_2\to\iota_3}(\cst{f}, \cst{a}), y) \noteq \cst{c}$ and
$\cst{@}_{\iota_2,\iota_3}(x, \cst{b}) \eq z$,
%
where $\cst{@}_{\tau,\upsilon}$ is a type-indexed family of
symbols representing the application of a function of type $\tau
\to \upsilon$.
\end{rep}%
With the applicative encoding, generation schemes can take the symbols
$\cst{@}_{\tau,\upsilon}$ into account, thereby exploiting the type
information carried by such symbols.\begin{rep} Since $\cst{@}_{\iota_2,\iota_3}$ is a
conjecture symbol, some weight generation scheme could give it a low weight,
which would also impact the second clause. By contrast, the native \lfhol{}
clauses share no symbols; the connection between them is hidden in the types
of variables and symbols that are ignored by the heuristics.

@@

The side conditions of the superposition rules SN and SP
(Sect.~\ref{sec:pre:rules}) rely on a literal
selection function to restrict the set of \emph{inference literals},
thereby reducing the search space. Given a clause, a literal selection
function returns a (possibly empty) subset of its literals. For
completeness, any nonempty subset selected must contain at least
one negative literal. If no literal is selected, all \emph{maximal}
literals become inference literals.
%Given a clause $L_1 \lor \cdots \lor L_n$, a literal selection function
%returns either $\{L_i\}$ for some $i$ or $\emptyset$.
The selection function E uses most often is probably
\texttt{SelectMaxLComplexAvoidPosPred}, which we abbreviate to
\texttt{SelectMLCAPP}. It selects at most one negative literal, based
on size, absence of variables, and maximality of the literal in the clause.
%It
%also avoids negative literals that share a predicate symbol with a
%positive literal in the same clause.

\begin{rep}
  Intuitively, applied variables can potentially be unified with more terms than
  terms with rigid heads. This makes them prolific in terms of possible
  inference partners, a behavior that can lead to creating many unnecessary
  clauses and should thus be avoided. On the other hand, shorter proofs might be
  found if we prefer selecting applied variables. To cover both scenarios, we
  implemented selection functions that prefer or defer selecting applied
  variables. % and that break ties using \texttt{SelectMLCAPP}.

@@

  Selection of the given clause is a critical choice point.
  %To ensure completeness every candidate clause should be selected
  %eventually\confrep{}{, unless it is identified as redundant and
  %discarded before its turn has come}. Ignoring a candidate forever
  %would compromise E's completeness guarantee.
  E heuristically assigns \emph{clause priorities} and \emph{clause weights} to
  the candidates.
  The priorities provide a crude partition, whereas the weights order the clauses
  within a partition.
  E's main loop visits a set of priority queues in round-robin fashion. From a given
  queue, the clause with the highest priority and the smallest weight is selected.
  \begin{rep}%
  Typically, one of the queues will use the clauses' age as priority, to ensure
  fairness (i.e., that each nonredundant clause is eventually chosen).
  \end{rep}

@@

}
Configurations that are suitable for a wide range of problems have emerged over
time. One of them is the configuration that is most often chosen by E's \emph{auto}
mode when running on TPTP benchmarks. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
%
%%% TYPESETTING: TEMPORARILY DISABLED LEFTMARGIN (?)
\begin{description}[labelwidth=\widthof{Precedence generation:}]
  \item[\rm Term order:] KBO
  \item[\rm Weight generation:] \texttt{invfreqrank}
  \item[\rm Precedence generation:] \texttt{invfreq}
  \item[\rm Literal selection:] \texttt{SelectMLCAPP}
  \item[\rm Clause selection:] ~ \\
\verb|1.CRSWeight(SimulateSOS, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.CRSWeight(ConstPrio, 0.1, 100, 100, 100, 100, 1.5, 1.5, 1.5),| \\
\verb|1.FIFOWeight(PreferProcessed),| \\
\verb|1.CRSWeight(PreferNonGoals, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.Refinedweight(SimulateSOS, 3, 2, 2, 1.5, 2)|
\end{description}
%
The clause selection scheme consists of five queues, each of which is
specified by a weight function template. The prefixes $n$\texttt{.}\ next to
the template names indicate that the queue will be visited $n$ times in the round-robin scheme
before moving to the next one. The first argument to each
template is the clause priority scheme.

@@

For first-order logic, skolemization preserves both satisfiability
(unprovability) and unsatisfiability (provability). In contrast, for
higher-order logics without the axiom of choice, naive skolemization is
unsound, because it introduces symbols that can be used to instantiate
higher-order variables.
%
One solution proposed by Miller \cite[Sect.~6]{dm-87-compact-proofs} is to ensure that
Skolem symbols are always applied to a minimal number of arguments.
%Accordingly, the version of \lfhol{} introduced by Bentkamp et al.\
%\cite{bbcw-21-lfho} distinguishes between mandatory and optional
%arguments.
However, to keep the implementation simple, we have decided to
ignore this issue and consider all arguments as optional, including those to
Skolem symbols. In Chapter \ref{ch:ehoh2} we extend Ehoh's logic to full higher-order logic
with the axiom of choice, which addresses the issue.
%TODO_PHD: Once Ehoh 2 paper is done, add reference to the section about choice here.

There is another transformation performed by preprocessing that is
problematic, but for a different reason. \emph{Definition unfolding} is the
process of replacing equationally defined symbols with their definitions and
removing the defining equations. A \relax{definition} is a clause of the form
$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
\{x_1,\dots,x_m\}$.
This transformation preserves unsatisfiability (provability) for first-~and
higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
is that by removing the definitional clause, we also remove a symbol $\cst{f}$
that otherwise could be used to instantiate a higher-order quantifier.
For example, the clause set
$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
higher-order logic, the second clause set would be unsatisfiable thanks to
the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-reduction.)
%
For the moment, we have simply disabled definition unfolding in Ehoh. We have
not measured the effect of this choice, but we conjecture it is not substantial.
Appropriately chosen term order will result in a similar outcome without
removing the defined symbol. We will enable definition unfolding again once we
have added support for $\lambda$-terms (Chapter \ref{ch:ehoh2}).

@@

The output may contain $\itrue$, $\ifalse$, or Bool\-ean
variables as nested formulas. Since E was first developed as an untyped prover,
unification of a variable with a Boolean constant was disallowed to avoid
unsoundness. We needed to undo this in Ehoh. Ehoh also removes trivial
literals $\ifalse \eq \itrue$ and $\itrue \noteq
\itrue$ that emerge during proof search.
\end{rep}

@@

%In this section, we consider the following questions:
How useful are Ehoh's new heuristics? And how does Ehoh perform compared with
E, used directly or in tandem with the
applicative encoding, and compared with other provers?
To answer the first question, we evaluated each new heuristic scheme independently. From
the empirical results, we derived a new configuration optimized for \lfhol{}.
For the second question, we compared Ehoh's success
rate\begin{rep} and speed\end{rep} on \lfhol{} problems with
native higher-order provers and
on applicatively encoded problems with E. We also included first-order
benchmarks to measure Ehoh's overhead.

@@

We used the \emph{boa} configuration as the basis to evaluate the new
heuristic schemes. For each heuristic parameter we tuned, we changed only its
value while keeping the other parameters the same as for \emph{boa}.
\begin{rep}This gave an idea of how each parameter affected overall
performance.\end{rep}
All heuristic parameters were tested on a 5012~problem suite generated
using Sledgehammer, consisting of four variants of the Judgment Day
\cite{bn-10-sh} suite.\begin{rep} The problems were given in native
\lfhol{} syntax.\end{rep} The experiments described in this subsection were
carried out using an earlier E version (2.3).\begin{conf}
%, exploiting its support for partial application and applied variables.
Our main findings are as follows:
%
\begin{itemize}
\item The combination of the weight generation scheme \texttt{invtypefreqrank}
  and the precedence generation scheme \texttt{invtypefreq} performs best.

@@

Clause selection is the heuristic component that we extended the most. We had to
assess the effect of a new heuristic weight function, a multiplier for the
occurrence of top-level applied variables, and clause priority based on the
number of top-level applied variables.

@@

To test the effect of the new type-based weight function, we replaced
\emph{boa}'s queue that uses
\texttt{4.CRS\-Weight(}\ldots\texttt{)}, with the queue ordered by
\texttt{4.CRSTypeWeight(}\ldots\texttt{)}. The original
heuristic is called \texttt{W} and the type-aware alternative \texttt{TW}.
%
We intuitively chose nine values for testing the effect of the applied
variable multiplier \textit{appv\_mult}.
%
Figure \ref{fig:ehoh:clauseweight-fun-eval} summarizes the results of combining
\texttt{W} or \texttt{TW} with the different
\textit{appv\_\allowbreak mult} values. Applying a multiplier smaller than
1, which corresponds to preferring literals containing applied variables, can
lose dozens of solutions. Overall, using the type-aware heuristic seems
slightly detrimental.
